A boundary-integral framework to simulate \\ the viscous erosion of a porous medium

### 2018-04-06 docs/methods/erosion_methods.tex

We validate our method on a single circular body placed symmetrically in
a channel flow.  By artificially fixing the area of the body as a function
of time, we are able to characterize the long-time behavior of the shape
and the resulting flow and these results agree with analytical results.
Then, we compute the vorticity of erosion in multiply-connected
geometries to better characterize erosion in porous media.

\paragraph{Limitations}
We make two fundamental design choices: first, we restrict ourselves to
two spatial dimensions, and second the bodies are considered immobile
(i.e.~fixed to their initial locations). The restriction to two
dimensions allows us to simulate the erosion of many bodies (up to $50$
in this manuscript) with high resolution, giving us the ability to
accurately characterize the erosion of a porous medium. While BIEs are
available in three dimensions, the extra computational cost limits the
number of bodies that could be practically simulated, making
porous-medium simulations less feasible. The second assumption of
immobile bodies affords a simple framework to study the effects of
erosion in isolation. While other effects, such as sedimentation or
transport, undoubtedly play a role in real applications, these effects
would introduce additional complexity, both conceptually and
numerically. We therefore defer the study of these effects to future
work.

\paragraph{Related Work}
The effect of erosion on a geometry is quite rich.  For instance, it has
been hypothesized and numerically observed that the erosion leads to a
shear stress that is uniform on the body which results in a uniform
erosion rate.  The shape that has a uniform shear stress is analyzed in
three dimensions by Montenegro-Johnson and Lauga~\cite{mon-lau2015}
which is an extension of the analysis done by Pironneau~\cite{pir1973}
to find a Stokesian drag minimizing shape. 
%\todo[inline]{Double check this statement. I am not sure that the shape found by Lauga has uniform shear stress. The shape found by Pironneau does.}
In this idealized case, analytical axisymmetric shapes that minimize the
drag can be found.  However, the effect of erosion on a porous media is
less understood.  The work most closely related to ours simulates erosion in a three-dimensional incompressible Stokesian fluid~\cite{mit-spa2016}, but they only consider single eroding body.

To simulate Stokesian erosion, it is essential to resolve the nonlocal
effects between the different bodies.  BIEs naturally resolve these
interactions, and they have been used extensively to simulate viscous
fluids in complex geometries, including porous media
flow~\cite{dea-qua-bir-jua2018, bar-mar-vee-zha2016}. Challenges of a
BIE formulation include resolving the interactions between nearly
touching bodies and developing efficient preconditioners for the
discretized system.  Possible preconditioners include
multigrid~\cite{qua-bir2015a} and the inverse fast multipole
method~\cite{qua-cou-dar2018, cou-pou-dar2017}, but the results
presented in this work require no more than $\mathcal{O}(100)$ GMRES
iterations which is too small to make preconditioners for BIEs
effective; preconditioners will be considered in future work when many
more grains are eroding and being transported by the flow.  To address
the nearly-singular integrands, we use the fact that the grains are
eroding and not moving, so their separation increases with time.
Therefore, in this work, we apply an upsampling scheme, but in future
work, we will require a near-singular integration
scheme~\cite{qua-bir2014a, klo-bar-gre-one2013, bar-wu-vee2015,
hel-oja2008a, bea-lai2001}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Formulation}
\label{s:formulation}
We start by defining the main variables used to model erosion.  We
consider flows that are confined by a solid wall $\Gamma$ that encloses
$M$ eroding bodies as diagrammed in Fig.~\ref{fig:schematic}.  The
bodies are denoted as $\gamma_\ell$, $\ell=1,\ldots,M$, and we write
$\gamma = \gamma_1 \cup \cdots \cup \gamma_M$.  On each body, $\nn$ and
$\ss$ denote the unit normal and unit tangent vectors respectively.
Throughout the paper, we adopt the convention that $\ss$ points in the
counterclockwise direction and $\nn$ points out of the fluid domain
(i.e.~into the bodies). Neglecting inertial forces, the dynamics of the
fluid is fully determined by the position of the bodies $\xx_\ell(s,t)
\in \gamma_\ell$, where $s$ is the arclength and $t$ is time.  Given
$\xx_\ell(s,t)$, $\ell=1,\ldots,M$, derived variables include the fluid
velocity $\uu$, the pressure $p$, and the shear stress $\tau$.   On the
bounding wall, we prescribe a velocity $\UU(\xx,t)$.  Then, the
governing equations are
\begin{equation}
\label{eqn:erosionModel}
\begin{split}
  \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap &&\mbox{conservation
of momentum}\\
\grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap
&&\mbox{\em conservation of mass} \\
\uu = 0, &\hspace{20pt} \xx \in \gamma, \gap &&\mbox{\em no slip on the
bodies} \\
\uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap &&\mbox{\em outer wall
velocity} \\
\Vn = \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
&&\mbox{\em erosion model},
\end{split}
\end{equation}
where $\Vn$ is the normal velocity of the interface and $\mu$ is the
fluid viscosity.  There are several models for
erosion~\cite{par-izu2000,lag2000}, but we have adopted the model that
erosion occurs in the normal direction at a rate that is proportional to
the magnitude of the shear stress~\cite{moo-ris-chi-zha-she2013}.  
\todo[inline]{This makes it sound the erosion law sound too will-nilly; like anyone can come up with their own erosion law tomorrow and it would be equally valid. This is not the case. We use the law in which erosion is proportional to shear stress because {\em that is what the laboratory experiments show is true} (Ristoph 2012).}

The constant of proportionality only sets the time scale, so we have set
$\Vn = |\tau|$ as described in equation~\eqref{eqn:erosionModel}.  For
the sake of numerical stability, we modify this law in
Section~\ref{sec:thetaL}.

@@

In this section, we describe an equivalent BIE formulation of the
incompressible Stokes equations with Dirichlet boundary conditions
\begin{equation}
  \begin{split}
  \mu\Delta \uu &= \grad p, \qquad && \xx \in \Omega, \\
  \grad \cdot \uu &= 0,   && \xx \in \Omega, \\
  \uu &= \ff,  && \xx \in \bd\Omega.
  \end{split}
  \label{eqn:stokes}
\end{equation}
The boundary condition $\ff$ is the prescribed velocity $\UU$ on
$\Gamma$, and it is zero on the eroding grains $\gamma$.  In two
dimensions, a BIE for a stream function formulation is
possible~\cite{gre-kro-may1996}, but this does not extend to three
dimensions.  Therefore, we adopt a primitive variable formulation where
the velocity is written in terms of the standard double-layer
potential~\cite{lad1963,poz1992}
\begin{align}
  \DD[\eeta](\xx) = \frac{1}{\pi} \int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2} 
    \eeta(\yy) \, ds_\yy, \quad \xx \in \Omega,
    \label{eqn:velocityDLP}
\end{align}
where $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\eeta$ is an unknown
density function.  As the viscosity $\mu$ also sets the time scale, we
take $\mu=1$.  \todo[inline]{Do you think we should flesh out this
time-scale stuff more precisely? i.e.~By setting $\mu = 1$ and $C = 1$
(the erosion constant), what exactly is the timescale that defines our
dimensionless time.} \todo[inline]{This might be worth doing.  Is it
something you can do quickly or do you want me to take a crack at it?}

>>>>

We supplement our primarily computational study with a few analytical results. In particular, we find a single body eroding in Stokes flow tends to a slender, for-aft symmetric form with sharp corners at its front and rear. By appealing to the principle that erosion more evenly distributes the shear stress along the surface, we can analytically predict the opening angle formed at the front and rear. In addition, drag laws from slender-body theory allows us to derive scaling laws for the rate of the body's area and drag reduction. These analytical results, in addition to having value in their own right, allow us to conclusively validate our simulation.

\paragraph{Limitations}
We make two fundamental design choices in our simulation: first, we restrict ourselves to
two spatial dimensions, and second the bodies are considered immobile
(i.e.~fixed to their initial locations). The restriction to two
dimensions allows us to simulate the erosion of many bodies (up to $50$
in this manuscript) with high resolution, giving us the ability to
accurately characterize the erosion of a porous medium. While BIEs are
available in three dimensions, the extra computational cost limits the
number of bodies that could be practically simulated, making
porous-medium simulations less feasible. The second assumption of
immobile bodies affords a simple framework to study the effects of
erosion in isolation. While other effects, such as sedimentation or
transport, undoubtedly play a role in real applications, these effects
would introduce additional complexity, both conceptually and
numerically. We therefore defer the study of these effects to future
work.

\paragraph{Related Work}

The most closely related work to ours is Mitchell \& Spagnolie \cite{mit-spa2016}, who developed methods to simulate viscous erosion in three dimensions. That work employed a traction integral equation to solve for the surrounding Stokes flow. The advantage to this formulation is that it immediately provides the shear stress required to evolve boundaries. However, the standard double-layer potential used here will allow us to compute certain quantities in the fluid bulk, such as vorticity and pressure, that will facilitate our understanding of the erosion process. Furthermore, our primary aim is to simulate the erosion of multiple bodies held fixed within a porous network, whereas Mitchell \& Spagnolie simulated a single eroding body that is mobile.

For these porous network simulations, it is essential to resolve the nonlocal effects between the different bodies.  BIEs naturally resolve these interactions, and they have been used extensively to simulate viscous fluids in complex geometries, including porous media flow~\cite{dea-qua-bir-jua2018, bar-mar-vee-zha2016}.  Common challenges of a BIE formulation include developing efficient preconditioners for the discretized system and numerical difficulties associated with nearly touching bodies. In this work, we consider geometries requiring $\mathcal{O}(100)$ GMRES iterations, so that preconditioners are not necessary. In future work, we will address mobile bodies for which near contacts may necessitate the preconditoners. Similarly, near-singular schemes will not be necessary due to the fact that the spacing between receding bodies of fixed position always grows in time. Again, the future consideration of mobile bodies may necessitate these near-singular schemes.

Possible preconditioners include
multigrid~\cite{qua-bir2015a} and the inverse fast multipole
method~\cite{qua-cou-dar2018, cou-pou-dar2017}, but the results
presented in this work require no more than $\mathcal{O}(100)$ GMRES
iterations which is too small to make preconditioners for BIEs
effective; preconditioners will be considered in future work when many
more grains are eroding and being transported by the flow.  
To address the nearly-singular integrands, we use the fact that the grains are
eroding and not moving, so their separation increases with time.
Therefore, in this work, we apply an upsampling scheme, but in future
work, we will require a near-singular integration
scheme~\cite{qua-bir2014a, klo-bar-gre-one2013, bar-wu-vee2015,
hel-oja2008a, bea-lai2001}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Formulation}
\label{s:formulation}
We start by defining the main variables used to model erosion.  We
consider flows that are confined by a solid wall $\Gamma$ that encloses
$M$ eroding bodies as diagrammed in Fig.~\ref{fig:schematic}.  The
bodies are denoted as $\gamma_\ell$, $\ell=1,\ldots,M$, and we write
$\gamma = \gamma_1 \cup \cdots \cup \gamma_M$.  On each body, $\nn$ and
$\ss$ denote the unit normal and unit tangent vectors respectively.
Throughout the paper, we adopt the convention that $\ss$ points in the
counterclockwise direction and $\nn$ points out of the fluid domain
(i.e.~into the bodies). Neglecting inertial forces, the dynamics of the
fluid is fully determined by the position of the bodies $\xx_\ell(s,t)
\in \gamma_\ell$, where $s$ is the arclength and $t$ is time.  Given
$\xx_\ell(s,t)$, $\ell=1,\ldots,M$, derived variables include the fluid
velocity $\uu$, the pressure $p$, and the shear stress $\tau$.   On the
bounding wall, we prescribe a velocity $\UU(\xx,t)$.  Then, the
governing equations are
\begin{equation}
\label{eqn:erosionModel}
\begin{split}
  \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap &&\mbox{conservation
of momentum}\\
\grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap
&&\mbox{\em conservation of mass} \\
\uu = 0, &\hspace{20pt} \xx \in \gamma, \gap &&\mbox{\em no slip on the
bodies} \\
\uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap &&\mbox{\em outer wall
velocity} \\
\Vn = \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
&&\mbox{\em erosion model},
\end{split}
\end{equation}
where $\Vn$ is the normal velocity of the interface and $\mu$ is the
fluid viscosity.   
The constant of proportionality only sets the time scale, so we have set
$\Vn = |\tau|$ as described in equation~\eqref{eqn:erosionModel}.  For
the sake of numerical stability, we modify this law in
Section~\ref{sec:thetaL}.

@@

In this section, we describe an equivalent BIE formulation of the
incompressible Stokes equations with Dirichlet boundary conditions
\begin{equation}
  \begin{split}
  \mu\Delta \uu &= \grad p, \qquad && \xx \in \Omega, \\
  \grad \cdot \uu &= 0,   && \xx \in \Omega, \\
  \uu &= \ff,  && \xx \in \bd\Omega.
  \end{split}
  \label{eqn:stokes}
\end{equation}
The boundary condition $\ff$ is the prescribed velocity $\UU$ on
$\Gamma$, and it is zero on the eroding grains $\gamma$.  In two
dimensions, a BIE for a stream function formulation is
possible~\cite{gre-kro-may1996}, but this does not extend to three
dimensions.  Therefore, we adopt a primitive variable formulation where
the velocity is written in terms of the standard double-layer
potential~\cite{lad1963,poz1992}
\begin{align}
  \DD[\eeta](\xx) = \frac{1}{\pi} \int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2} 
    \eeta(\yy) \, ds_\yy, \quad \xx \in \Omega,
    \label{eqn:velocityDLP}
\end{align}
where $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\eeta$ is an unknown
density function.  As the viscosity $\mu$ also sets the time scale, we
take $\mu=1$.
